---
title: "Predictive Analytics: Problem Sets 7"
author: "Drohanash Majumdar, 704"
date: "2026-03-23"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

if (!require(ISLR)) install.packages("ISLR")
if (!require(leaps)) install.packages("leaps")
if (!require(glmnet)) install.packages("glmnet")

library(ISLR)
library(leaps)
library(glmnet)
```

# Shrinkage Methods : Comparison of least squares, ridge regression, LASSO and Elastic Net

You would be working with Hitters data set available in the ISLR library. You have to predict the Salary based on all available variables. First make sure that you have deleted all rows with one or more missing values.
```{r ps7-setup}
data(Hitters)
Hitters <- na.omit(Hitters)
```

1. Split the data set into a training set and a test set randomly keeping 100 observations in test set and remaining in the training set.
```{r}
set.seed(123)
test_idx  = sample(1:nrow(Hitters), 100)
train7    = Hitters[-test_idx, ]
test7     = Hitters[test_idx, ]

X_train = model.matrix(Salary ~ ., train7)[, -1]
y_train = train7$Salary
X_test  = model.matrix(Salary ~ ., test7)[, -1]
y_test  = test7$Salary

cat("Train size:", nrow(train7), "| Test size:", nrow(test7), "\n")
```

2. Train the model using least squares and report the train MSE and test MSE.
```{r ps7-ols}
ols_fit   = lm(Salary ~ ., data = train7)
ols_train = predict(ols_fit, train7)
ols_test = predict(ols_fit, test7)

ols_train_mse = mean((y_train - ols_train)^2)
ols_test_mse = mean((y_test  - ols_test)^2)

cat("OLS Train MSE:", round(ols_train_mse, 2), "\n")
cat("OLS Test MSE: ", round(ols_test_mse,  2), "\n")
```

3. Fit ridge regression models on the training set, with grid values of λ. Report the following: 

(i) the dimension of the coefficient matrix so obtained.

(ii) The l2 norm of the regression coefficients corresponding to the 50th and 60th choice of λ. Comment on the output.

(iii) Test MSE corresponding to these two choices of λ.
```{r ps7-ridge}
# Grid of lambda values
lambda_grid = 10^seq(5, -2, length = 100)

ridge_fit = glmnet(X_train, y_train, alpha = 0, lambda = lambda_grid)

# (i) Dimension of coefficient matrix
cat("Coefficient matrix dimensions:", dim(coef(ridge_fit)), "\n")

# (ii) L2 norm at 50th and 60th lambda
l2_50 = sqrt(sum(coef(ridge_fit)[-1, 50]^2))
l2_60 = sqrt(sum(coef(ridge_fit)[-1, 60]^2))
cat("L2 norm at lambda[50] =", round(lambda_grid[50], 4), ":", round(l2_50, 4), "\n")
cat("L2 norm at lambda[60] =", round(lambda_grid[60], 4), ":", round(l2_60, 4), "\n")
cat("Larger lambda => smaller L2 norm (more shrinkage)\n")

# (iv) Test MSE for 50th and 60th lambda
pred_50 = predict(ridge_fit, s = lambda_grid[50], newx = X_test)
pred_60 = predict(ridge_fit, s = lambda_grid[60], newx = X_test)
cat("Test MSE at lambda[50]:", round(mean((y_test - pred_50)^2), 2), "\n")
cat("Test MSE at lambda[60]:", round(mean((y_test - pred_60)^2), 2), "\n")
```

4. Make a line plot (for all the predictors) of the standardized coefficients against the grid value of λ chosen by you. Interpret the graph.
```{r}
beta = as.matrix(ridge_fit$beta)
lambda = ridge_fit$lambda

matplot(lambda, t(beta), type = "l", lty = 1,
        xlab = "Lambda", ylab = "Coefficients",xlim = c(1e-2, 1e4))
```

**Interpretation:** As λ increases, all coefficients shrink toward zero. No coefficient reaches exactly zero — Ridge retains all predictors but with reduced magnitudes.

5. How would the graph in (4) change if instead of λ you plot on the x axis the ratio of l2 norm of the ridge regression coefficient to the least square regression coefficient.
```{r}
ridge_coefs = predict(ridge_fit, type = "coefficients")

# Compute L2 norm for each lambda
ridge_l2_norm = apply(ridge_coefs[-1, ], 2, function(b) sqrt(sum(b^2)))

# Least squares coefficients
lm_model = lm(y_train ~ X_train)
lm_coefs = coef(lm_model)[-1]
lm_l2_norm = sqrt(sum(lm_coefs^2))

# Ratio
ratio = ridge_l2_norm / lm_l2_norm

# Plot coefficients vs ratio
matplot(ratio, t(ridge_coefs[-1, ]), type = "l", lty = 1,
        xlab = "L2 Norm Ratio", ylab = "Coefficients")
```

6. Run the Ridge regression on training data by choosing λ by cross-validation. Find the test MSE for the optimal choice of λ obtained using cross validation.
```{r}
set.seed(42)
cv_ridge = cv.glmnet(X_train, y_train, alpha = 0)
best_lambda_ridge = cv_ridge$lambda.min
cat("Best lambda (CV):", round(best_lambda_ridge, 4), "\n")

ridge_pred_cv = predict(cv_ridge, s = best_lambda_ridge, newx = X_test)
ridge_test_mse = mean((y_test - ridge_pred_cv)^2)
cat("Ridge Test MSE (CV lambda):", round(ridge_test_mse, 2), "\n")
```

7. Fit a LASSO model on the training set, with λ chosen by cross-validation. Report the test MSE, along with the number of non-zero coefficient estimates.
```{r}
set.seed(42)
cv_lasso = cv.glmnet(X_train, y_train, alpha = 1)
best_lambda_lasso = cv_lasso$lambda.min
cat("Best lambda (CV):", round(best_lambda_lasso, 4), "\n")

lasso_pred = predict(cv_lasso, s = best_lambda_lasso, newx = X_test)
lasso_test_mse = mean((y_test - lasso_pred)^2)

lasso_coefs = coef(cv_lasso, s = best_lambda_lasso)
nonzero = sum(lasso_coefs != 0) - 1  # exclude intercept
cat("LASSO Test MSE:", round(lasso_test_mse, 2), "\n")
cat("Number of non-zero coefficients:", nonzero, "\n")
```

8. Fit an Elastic Net model on the training set, with λ1 and λ2 chosen by cross-validation. Report the test MSE. Also report the predictors with non-zero coefficients.
```{r}
alphas = seq(0.1, 0.9, by = 0.1)
cv_errors = sapply(alphas, function(a) {
  cv_fit = cv.glmnet(X_train, y_train, alpha = a)
  min(cv_fit$cvm)
})

best_alpha = alphas[which.min(cv_errors)]
cat("Best alpha:", best_alpha, "\n")

set.seed(42)
cv_enet     = cv.glmnet(X_train, y_train, alpha = best_alpha)
best_lam_en = cv_enet$lambda.min
cat("Best lambda (CV):", round(best_lam_en, 4), "\n")

enet_pred = predict(cv_enet, s = best_lam_en, newx = X_test)
enet_test_mse = mean((y_test - enet_pred)^2)

enet_coefs = coef(cv_enet, s = best_lam_en)
nonzero_en = sum(enet_coefs != 0) - 1
cat("Elastic Net Test MSE:", round(enet_test_mse, 2), "\n")
cat("Non-zero predictors:", nonzero_en, "\n")
cat("Non-zero predictor names:\n")
print(rownames(enet_coefs)[enet_coefs[,1] != 0 & rownames(enet_coefs) != "(Intercept)"])
```

9. Compare the test MSE ’s of least squares regression, Ridge regression, LASSO and ELASTIC NET and comment on how accurately we can predict the Salary of a baseball player.
```{r}
comparison = data.frame(
  Method     = c("OLS", "Ridge (CV)", "LASSO (CV)", "Elastic Net (CV)"),
  Test_MSE   = round(c(ols_test_mse, ridge_test_mse, lasso_test_mse, enet_test_mse), 2)
)
print(comparison)
```

**Interpretation:** Shrinkage methods (Ridge, LASSO, Elastic Net) typically outperform OLS on test data because they reduce overfitting. LASSO also performs variable selection (many zero coefficients), making the model more interpretable. Elastic Net combines Ridge and LASSO penalties and can outperform either alone. The method with the lowest test MSE gives the most accurate salary predictions.





